Temporal ordering and registration of images in studies of developmental dynamics

### 2014-06-05 image_analysis_paper/paper_figures2/fig5_2.tex
Added edits to paper
Added comments from Amit

%\usepackage[margin=0in, paperwidth=8.7cm, paperheight=9.5cm]{geometry}
\usepackage[margin=0in, paperwidth=8.7cm, paperheight=17cm]{geometry}

@@

\centering
\begin{tikzpicture}
	\node at (0cm, 0cm) {\includegraphics[width=6cm, angle=90]{data3_embed_bw}};
	\node at (-4cm, 2.5cm) {\textsf{A}};
\end{tikzpicture}\\
\begin{tikzpicture}
	\node at (0cm, 0cm) {\includegraphics[width=8cm]{data3_embed_color}};
	\node at (-4cm, 2.5cm) {\textsf{B}};
	%\draw[help lines,step=.5] (-4,-4) grid (4,4);
	\draw [draw=black!20!green, ->, ultra thick] (-2, -2) .. controls (-1.1, 0.2) and  (1, 0)  .. (2, -2);
	\draw [blue, ->, ultra thick] (-2.65, -1.8)  .. controls (-2.3, 0.9) and (-1.7, 1.5) .. (-1.4, 2.1); 
\end{tikzpicture}\\
\begin{tikzpicture}
	\node at (-4.25cm, 0cm) {\textsf{C}};
	\node at (0,0) {\includegraphics[width=8cm]{wt_trajectory}};
	%\draw[help lines,step=.5] (-4,-1) grid (4,1);
	\draw[->, color=white, thick] (-1.1cm, -0.1cm) -- (-1.3cm, -0.3cm);
	\draw[->, color=white, thick] (-0.7cm, -0.1cm) -- (-0.5cm, -0.2cm);

@@

\end{tikzpicture}\\
\begin{tikzpicture}
	\node at (-4.25cm, 0cm) {\textsf{D}};
	\node at (0,0) {\includegraphics[width=8cm]{mut_trajectory}};

>>>>

%\usepackage[margin=0in, paperwidth=8.7cm, paperheight=9.5cm]{geometry}
\usepackage[margin=0in, paperwidth=8.7cm, paperheight=11.7cm]{geometry}

@@

\centering
\begin{tikzpicture}
	\node at (0cm, 0cm) {\includegraphics[width=6.9cm, angle=90]{data3_embed_bw}};
	\node at (-4.35cm, 2.75cm) {\textsf{A}};
	%\draw[help lines,step=.5] (-4,-4) grid (4,4);
	\draw [draw=white!50!gray, ->, ultra thick] (-2.2, -2) .. controls (-1.1, 0.5) and  (1, 0)  .. (2, -2);
	\draw [draw=white!50!gray, ->, ultra thick] (-3, -1.8)  .. controls (-2.6, 0.9) and (-1.9, 1.5) .. (-1.7, 2.1); 
	\node at (2.05cm, 1.6cm) {\includegraphics[width=4.1cm]{data3_embed_color}};
\end{tikzpicture}\\
\begin{tikzpicture}
	\node at (-4.25cm, 0cm) {\textsf{B}};
	\node at (0,0) {\includegraphics[width=8cm]{wt_trajectory}};
	%\draw[help lines,step=.5] (-4,-1) grid (4,1);
	\draw[->, color=white, thick] (-1.1cm, -0.1cm) -- (-1.3cm, -0.3cm);
	\draw[->, color=white, thick] (-0.7cm, -0.1cm) -- (-0.5cm, -0.2cm);

@@

\end{tikzpicture}\\
\begin{tikzpicture}
	\node at (-4.25cm, 0cm) {\textsf{C}};
	\node at (0,0) {\includegraphics[width=8cm]{mut_trajectory}};

### 2014-06-05 image_analysis_paper/pnas_draft2.tex
Added comments from Amit

Image registration has been studied in a variety of contexts \cite{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}, 
and most registration algorithms are based on landmarks, such as the head and fins in \fig~\ref{fig:fish}.
%
In contrast, our approach requires no such information. 
%
It is based on a manifold learning algorithm (vector diffusion maps, VDM \cite{singer2012vector}) which simultaneously addresses the problems of registration and temporal ordering. 
%
This algorithm is one of several influential nonlinear dimensionality reduction techniques that have been developed over the past decade \cite{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear}. 
%
The technique has been used for data mining in
applications ranging from analysis of cryo-electron microscopy (cryo-EM) images of individual molecules to face recognition, and is sufficiently general to be applicable to a wide variety of biological imaging studies. 
%
Here the algorithm is adapted for simultaneous registration and temporal ordering of images from cross-sectional studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories.
%
To illustrate our approach, we analyze three data sets from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of developmental dynamics \cite{jaeger2012drosophila}.
%
%The first data set, shown in Figure~\ref{subfig:raw_data1}, is relatively simple and will allow us to both illustrate and validate this approach.
%
%In the second data set, shown in Figure~\ref{subfig:raw_data2}, the dynamics are significantly more complex.
%
%Here, we will show that we can recover a trajectory that is qualitatively consistent with current knowledge.
%
%The third set consists of data collected from both wild type and mutant embryos, where joint processing of the data will help to visualize the different sources of variability along distinct developmental paths.

@@

%%%YGK  when I read this I think of Eiffel tower pictures, not of ours - there is a transition... careful where we do it.
%
%Our goal is to organize data sets from cross-sectional studies of developmental processes in a meaningful and informative way.
%
%We will show that dimensionality reduction algorithms can simultaneously register and temporally order the images to give us a meaningful picture of the underlying developmental dynamics.
%
Our approach is based on vector diffusion maps \cite{singer2012vector}, a manifold learning
technique developed for data sets which contain two types of sources of variability:
geometric symmetries (such as translations and rotations of the images due to the experimental setup) which one would like to factor out,
and ``additional" directions of variability (such as temporal dynamics) which one wants to uncover.
%
Vector diffusion maps combine two algorithms, {\em angular synchronization} \cite{singer2011angular} for image registration and {\em diffusion maps} \cite{coifman2005geometric} for extracting intrinsic low-dimensional structure in data, into a single computation that allows us to simultaneously register and temporally order our images.
%
%Vector diffusion maps has been used to classify images from cryo-electron microscopy (cryo-EM) experiments, where each image is a projection of a randomly oriented molecule.
%
%%%YGK to imply which rotations we need to have discussed it above
%%
We will use the algorithm to register images of {\it Drosophila} embryos with respect to rotations and translations, as well as uncover the main direction of variability {\it after} removing translational and rotational symmetries.
%
We assume that, in these sets of images, the main direction of variability is parameterized by the developmental time of each embryo, so that uncovering this direction will allow us to recover the developmental dynamics.

@@

As a first illustration of using vector diffusion maps to temporally order and register images, we analyze a data set where the correct registration and temporal order is known {\it a priori}.
%
This data set is obtained by fluorescent imaging of {\it Drosophila} embryos during the third hour after fertilization.
%
By this stage of development, sequential divisions of the fertilized egg have generated a system where most of the nuclei are arranged in a monolayer under the common plasma membrane \cite{foe1983studies}.
%
\fig~\ref{subfig:raw_data1} shows optical cross-sections of each embryo, with nuclei (gray in the figure) labeled by DAPI, a DNA stain.
%
During the third hour of development, this arrangement of nuclei is patterned by the concentration profiles of chemical signals that control cell differentiation.
%
One of these signals is provided by the nuclear localization of the transcription factor Dorsal (Dl, green in the figure), which subdivides the early embryo into the territories that give rise to the muscle, nerve, and skin tissues.
%
Another signal is provided by the phosphorylated form of the extracellular signal regulated kinase (dpERK, red in the figure), an enzyme that specifies a subset of neuronal cells.
%

@@

Visually, the dynamics in this data set are much more complicated, and only 22\% of the variability in the images is captured by the first principal component.
%
Simple time-ordering based on this linear projection is therefore not possible (see \fig~\ref{fig:PCA_data2}).
%
%Even when projecting on several leading principal components (see SI), a nonlinear approach, such as vector diffusion maps, is required to temporally order the data.
%
\fig~\ref{subfig:ordered_data2} shows the images registered and ordered using vector diffusion maps \cite{singer2012vector}, 
and visual inspection confirms this ordering is consistent with the known developmental dynamics outlined above.
%
However, assimilating the dynamics even from this concatenation of images is difficult for two reasons, even after registration and ordering.
%
First, the sheer number of images makes visual processing of the entire data set nontrivial;
and second, the ordered data is not entirely smooth due to inter-embryo variability.
%
To highlight the developmental dynamics revealed by vector diffusion maps, we applied a Gaussian filter to the trajectory of registered and ordered images to obtain an averaged developmental path (see also \cite{kemelmacher2011exploring}).  
%
Sequential snapshots of this averaged trajectory, shown in \fig~\ref{subfig:averaged_data2}, serve as a summary of the stereotypic developmental dynamics reconstructed from cross-sectional data.
%
Thus, vector diffusion maps can accomplish the image registration and ordering tasks presented in the caricature in \fig~\ref{fig:fish}, even in the absence of information about image landmarks (e.g., fins) and without {\it a priori} knowledge of developmental dynamics (e.g., correlation of age with body size).

@@

We presented a unified approach to temporal ordering and registration of images in cross-sectional studies of developmental dynamics. 
%
To the best of our knowledge, algorithmic approaches to these two tasks have been explored largely independently of each other. 
%
In particular, temporal ordering of imaging data sets was done with a significant amount of human supervision and using registered images as a starting point \cite{yuan2014automated, surkova2008characterization}.  
%
In parallel, ordering of large-scale cross-sectional data was done in the context of molecular profiling studies, in which data are vectors describing the amounts of different chemical species, such as mRNA abundances \cite{anavy2014blind, trapnell2014dynamics, gupta2008extracting}. 
%
Automated temporal ordering of such data sets was accomplished by first projecting the data on a low-dimensional subspace spanned by the leading principal or independent components, and then solving a traveling salesman problem or constructing a minimum spanning tree on the projected data to order multiple snapshots. 
%
The approach presented here is different because it addresses image registration and ordering in a single step and orders the data without first projecting them on a low-dimensional subspace. 
%
The task of image registration has also been widely studied \cite{zitova2003image}, for applications such as face recognition \cite{rowley1998rotation}, medical image registration \cite{hajnal2010medical}, and texture classification \cite{greenspan1994rotation}.
%
In contrast to most of the existing approaches to image registration which rely on the knowledge about some landmarks in the data \cite{ian1998statistical} (such as the eyes in face recognition applications \cite{zhao2003face}), algorithms based on angular synchronization (such as vector diffusion maps) can register images even in the absence of such information. 
%TODO: look at Eisen Cell paper about registration

@@

Previously, angular synchronization and vector diffusion maps have been used to reconstruct molecular shapes from cryo-electron microscopy imaging data \cite{singer2012vector, singer2011three}. 
%
Because of the high levels of instrument noise in these images, thousands of images were needed for successful shape reconstruction. 
%
Based on our work, we expect that much smaller data sets may be sufficient for successful reconstruction of developmental trajectories, providing the first step in systems level studies of embryogenesis. 
%
In the future, functional data analysis tools can be used for the quantitative estimation of both the average trajectory and its variability \cite{silverman2005functional}.  
%
One can also test which physical variable, among a set of candidate variables, is most correlated with the uncovered vector diffusion maps coordinate to gain insight into which physical quantities change across the developmental trajectory (e.g., total intensity of one signal, roundness of the embryo).
%
Here, we have used vector diffusion maps to register the images with respect to translations and rotations; however, it is conceivable that one could factor out other sources of variability, such as reflections \cite{singer2012vector, goemans1995improved, bandeira2013cheeger} or dilations, in the images.
%
One might even envision automated analysis of embryos imaged in arbitrary three-dimensional orientations, in the spirit of cryo-EM image analysis.
%
We believe that algorithms based on image features which are insensitive to small deformations, such as the scattering transform \cite{mallat2012group}, are especially promising.
%
Another important application is related to the joint analysis of data sets provided by different imaging approaches, such as merging live imaging data of tissue morphogenesis with snapshots of cell signaling and gene expression from fixed embryos \cite{krzic2012multiview, ichikawa2014live, rubel2010coupling}. 
%
Given the rapidly increasing volumes of imaging data from studies of multiple developmental systems, we expect that dimensionality reduction approaches discussed in this work will be increasingly useful for biologists and motivate future applications and algorithmic advances.

@@

\begin{acknowledgments}
The authors thank Adam Finkelstein,  Thomas Funkhouser, and John Storey for helpful discussions. 
%
C.J.D. was supported by the Department of Energy Computational Science Graduate Fellowship (CSGF), grant number DE-FG02-97ER25308.
%
B.L. and S.Y.S. were supported by the National Institutes of Health Grant R01GM086537. 
%
T.J.L. and H.L. were supported by the National Science Foundation Grant Emerging Frontiers in Research and Innovation (EFRI) 1136913.
%
I.G.K. was supported by the National Science Foundation (CS\&E program).
\end{acknowledgments}

@@

\bibitem{singer2011three}
Singer A, Shkolnisky Y (2011) Three-dimensional structure determination from
  common lines in cryo-em by eigenvectors and semidefinite programming.
  \textit{SIAM J Imaging Sci} 4(2):543--572.

>>>>

Image registration has been studied in a variety of contexts \cite{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}, 
and most registration algorithms are based on landmarks, such as the head and fins in \fig~\ref{fig:fish}.
%
In contrast, our approach requires no such information. 
%
It is based on a manifold learning algorithm (vector diffusion maps, VDM \cite{singer2012vector}) which simultaneously addresses the problems of registration and temporal ordering. 
%
This algorithm is one of several influential nonlinear dimensionality reduction techniques that have been developed over the past decade \cite{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear}. 
%
The technique has been used for data mining in
applications ranging from analysis of cryo-electron microscopy (cryo-EM) images of individual molecules to face recognition, and is sufficiently general to be applicable to a wide variety of biological imaging studies {\color{red} does this sentence need references?}. 
%
Here the algorithm is adapted for the analysis of images from cross-sectional studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories.
%
To illustrate our approach, we analyze three data sets from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of developmental dynamics \cite{jaeger2012drosophila}.
%
%The first data set, shown in Figure~\ref{subfig:raw_data1}, is relatively simple and will allow us to both illustrate and validate this approach.
%
%In the second data set, shown in Figure~\ref{subfig:raw_data2}, the dynamics are significantly more complex.
%
%Here, we will show that we can recover a trajectory that is qualitatively consistent with current knowledge.
%
%The third set consists of data collected from both wild type and mutant embryos, where joint processing of the data will help to visualize the different sources of variability along distinct developmental paths.

@@

%%%YGK  when I read this I think of Eiffel tower pictures, not of ours - there is a transition... careful where we do it.
%
%Our goal is to organize data sets from cross-sectional studies of developmental processes in a meaningful and informative way.
%
%We will show that dimensionality reduction algorithms can simultaneously register and temporally order the images to give us a meaningful picture of the underlying developmental dynamics.
%
Our approach is based on vector diffusion maps \cite{singer2012vector}, a manifold learning
technique developed for data sets which contain two types of sources of variability:
geometric symmetries (such as translations and rotations of the images due to the experimental setup) which one would like to factor out,
and ``additional" directions of variability (such as temporal dynamics) which one wants to uncover.
%
Vector diffusion maps combine two algorithms, {\em angular synchronization} \cite{singer2011angular} for image registration and {\em diffusion maps} \cite{coifman2005geometric} for extracting intrinsic low-dimensional structure in data, into a single computation.
%
%Vector diffusion maps has been used to classify images from cryo-electron microscopy (cryo-EM) experiments, where each image is a projection of a randomly oriented molecule.
%
%%%YGK to imply which rotations we need to have discussed it above
%%
We will use the algorithm to register images of {\it Drosophila} embryos with respect to rotations and translations, as well as uncover the main direction of variability {\it after} removing translational and rotational symmetries.
%
We assume that, in these sets of images, the main direction of variability is parameterized by the developmental time of each embryo, so that uncovering this direction will allow us to recover the developmental dynamics.

@@

As a first illustration, we analyze a data set where the correct registration and temporal order is known {\it a priori}.
%
This data set is obtained by fluorescent imaging of {\it Drosophila} embryos during the third hour after fertilization.
%
By this stage of development, sequential divisions of the fertilized egg have generated a system where most of the nuclei are arranged in a monolayer under the common plasma membrane \cite{foe1983studies}.
%
\fig~\ref{subfig:raw_data1} shows optical cross-sections of each embryo, with nuclei (gray in the figure) labeled by DAPI, a DNA stain.
%
During the third hour of development, this arrangement of nuclei is patterned by the concentration profiles of chemical signals that control cell differentiation.
%
One of these signals is provided by the nuclear localization of the transcription factor Dorsal (Dl, green in the figure), which subdivides the early embryo into the territories that give rise to the muscle, nerve, and skin tissues.
%
Another signal is provided by the phosphorylated form of the extracellular signal regulated kinase (dpERK, red in the figure), an enzyme that specifies a subset of neuronal cells.
%

@@

Visually, the dynamics in this data set are much more complicated, and only 22\% of the variability in the images is captured by the first principal component.
%
Simple time-ordering based on this linear projection is therefore not possible (see \fig~\ref{fig:PCA_data2}).
%
%Even when projecting on several leading principal components (see SI), a nonlinear approach, such as vector diffusion maps, is required to temporally order the data.
%
\fig~\ref{subfig:ordered_data2} shows the images registered and ordered using vector diffusion maps \cite{singer2012vector}, 
and visual inspection confirms this ordering is consistent with the known developmental dynamics outlined above.
%
However, assimilating the dynamics even from this concatenation of images is difficult for two reasons, even after registration and ordering.
%
First, the sheer number of images makes visual processing of the entire data set nontrivial;
and second, the ordered data is not entirely smooth due to inter-embryo variability.
%
To highlight the developmental dynamics revealed by vector diffusion maps, we applied a Gaussian filter to the trajectory of registered and ordered images to obtain an averaged developmental path (see also \cite{kemelmacher2011exploring}).  
%
Sequential snapshots of this averaged trajectory, shown in \fig~\ref{subfig:averaged_data2}, serve as a summary of the stereotypic developmental dynamics reconstructed from cross-sectional data.
%
Thus, vector diffusion maps can accomplish the tasks presented in the caricature in \fig~\ref{fig:fish}, even in the absence of information about image landmarks (e.g., fins) and without {\it a priori} knowledge of developmental dynamics (e.g., correlation of age with body size).

@@

We presented a unified approach to temporal ordering and registration of images in cross-sectional studies of developmental dynamics. 
%
To the best of our knowledge, algorithmic approaches to these two tasks have been explored largely independently of each other. 
%
In particular, temporal ordering of imaging data sets was done with a significant amount of human supervision and using registered images as a starting point \cite{yuan2014automated, surkova2008characterization}.  
%
In parallel, ordering of large-scale cross-sectional data was done in the context of molecular profiling studies, in which data are vectors describing the amounts of different chemical species, such as mRNA abundances \cite{anavy2014blind, trapnell2014dynamics, gupta2008extracting}. 
%
Automated temporal ordering of such data sets was accomplished by first projecting the data on a low-dimensional subspace spanned by the leading principal or independent components, and then solving a traveling salesman problem or constructing a minimum spanning tree on the projected data to order multiple snapshots. 
%
The approach presented here is different because it addresses image registration and ordering in a single step and orders the data without first projecting them on a low-dimensional subspace. 
%
The task of image registration has also been widely studied \cite{zitova2003image}, for applications such as face recognition \cite{rowley1998rotation}, medical image registration \cite{hajnal2010medical}, and texture classification \cite{greenspan1994rotation}.
%
In contrast to most of the existing approaches to image registration which rely on the knowledge about some landmarks in the data \cite{ian1998statistical} (such as the eyes in face recognition applications \cite{zhao2003face}), algorithms based on angular synchronization (such as vector diffusion maps) can register images even in the absence of such information.

%TODO: look at Eisen Cell paper about registration

@@

{\color{red} Previously, angular synchronization and vector diffusion maps have been used to reconstruct molecular shapes from cryo-electron microscopy imaging data \cite{singer2012vector, zhao2014rotationally, singer2011viewing}.} 
%
Because of the high levels of instrument noise in these images, thousands of images were needed for successful shape reconstruction. 
%
Based on our work, we expect that much smaller data sets may be sufficient for successful reconstruction of developmental trajectories, providing the first step in systems level studies of embryogenesis. 
%
In the future, functional data analysis tools can be used for the quantitative estimation of both the average trajectory and its variability \cite{silverman2005functional}.  
%
One can also test which physical variable, among a set of candidate variables, is most correlated with the uncovered vector diffusion maps coordinate to gain insight into which physical quantities change across the developmental trajectory (e.g., total intensity of one signal, roundness of the embryo).
%
Here, we have used vector diffusion maps to register the images with respect to translations and rotations; however, it is conceivable that one could factor out other sources of variability, such as reflections \cite{singer2012vector, goemans1995improved, bandeira2013cheeger} or dilations, in the images.
%
One might even envision automated analysis of embryos imaged in arbitrary three-dimensional orientations, in the spirit of cryo-EM image analysis.
%
{\color{red} In these situations, vector diffusion maps is essential, as global registration is topologically impossible \cite{zhao2014rotationally}.}
%
{\color{red} We believe that algorithms based on rotation- and translation-invariant image features which are stable to small deformations, such as the scattering transform \cite{mallat2012group} or steerable PCA and the bispectrum \cite{zhao2014rotationally}, are especially promising; such methods circumvent the need for registration by using these invariant features in the data analysis.}
%
Another important application is related to the joint analysis of data sets provided by different imaging approaches, such as merging live imaging data of tissue morphogenesis with snapshots of cell signaling and gene expression from fixed embryos \cite{krzic2012multiview, ichikawa2014live, rubel2010coupling}. 
%
Given the rapidly increasing volumes of imaging data from studies of multiple developmental systems, we expect that dimensionality reduction approaches discussed in this work will be increasingly useful for biologists and motivate future applications and algorithmic advances.

@@

\begin{acknowledgments}
The authors thank Adam Finkelstein,  Thomas Funkhouser, and John Storey for helpful discussions. 
%
C.J.D. was supported by the Department of Energy Computational Science Graduate Fellowship (CSGF), grant number DE-FG02-97ER25308.
%
B.L. and S.Y.S. were supported by the National Institutes of Health Grant R01GM086537. 
%
T.J.L. and H.L. were supported by the National Science Foundation Grant Emerging Frontiers in Research and Innovation (EFRI) 1136913.
%
A.S. was supported by the Air Force Office of Scientific Research Grant
FA9550-12-1-0317.
%
I.G.K. was supported by the National Science Foundation (CS\&E program).
\end{acknowledgments}

@@

\bibitem{zhao2014rotationally}
Zhao Z, Singer A (2014) Rotationally invariant image representation for
viewing direction classification in cryo-EM. \textit{J Struct Biol} 186(1):153--166.

\bibitem{singer2011viewing}
Singer A, Zhao Z, Shkolnisky Y, Hadani R (2011) Viewing angle
classification of cryo-electron microscopy images using eigenvectors. \textit{SIAM J Imaging Sci} 4(2):723--759.

%\bibitem{singer2011three}
%Singer A, Shkolnisky Y (2011) Three-dimensional structure determination from
%  common lines in cryo-em by eigenvectors and semidefinite programming.
%  \textit{SIAM J Imaging Sci} 4(2):543--572.
